Mamba-Based Tumor Segmentation\\ \thanks{$\dagger$ Equal contributions}

### 2024-04-20 paper/main.tex
paper

\section{Results}
We now compare our model against three other common segmentation models, U-Mamba \cite{}, U-Net \cite{}, and SegViT \cite{}. And it was good.

\begin{table}[!h]
    \renewcommand{\arraystretch}{1.2}
    \setlength{\tabcolsep}{2pt}
    \centering
    \footnotesize
    \begin{tabular}{|l|c|c|c|c|}
    \hline
    Model & Params & Size & IOU & Dice \\
    \hline
    Pyramidal Mamba (Ours) & 10M & 109.15 & 0.8004 & 0.8867 \\
    \hline
    U-Mamba & 9.8M & 108.10 & 0.7908 & 0.8799 \\
    \hline
    U-Net & 7.8M & 31.09 & 0.8071 & 0.8929 \\
    \hline
    SegViT & Y & Z & A & B \\
    \hline
    ResNet18 & 13.9M & 55.76 & 0.7623 & 0.8648 \\
    \hline
    ResNet50 & 42.9M & 171.85 & 0.7661 & 0.8672 \\
    \hline
    Ultra LightM-UNet & 0.049M & 0.20 & 0.6725 & 0.7955 \\
    \hline
    \end{tabular}
    \end{table}

@@

\bibitem{dosovitskiy2021image} A. Dosovitskiy et al., "An Image is Worth 16x16 Words: Transformers for Image Recognition at Scale," arXiv:2010.11929 [cs.CV], 2021.
\end{thebibliography}
\end{document}

>>>>

\subsection{UltraLight VM-UNet}
The UltraLight VM-UNet \cite{ultralightvmunet} is a lightweight neural network architecture designed for skin lesion segmentation tasks. It is built upon the Vision Mamba module, which is a state-space model (SSM) that can efficiently handle long-range dependencies in sequences, making it well-suited for image segmentation tasks.

The key innovation of the UltraLight VM-UNet is the proposed Parallel Vision Mamba Layer (PVM Layer), which processes deep features in parallel using multiple Vision Mamba blocks. Specifically, the input feature map is split into multiple sub-feature maps, each processed by a separate Vision Mamba block with a reduced channel count. This parallel processing approach allows the UltraLight VM-UNet to maintain high segmentation performance while significantly reducing the number of parameters and computational complexity.

The UltraLight VM-UNet is reported to have only 0.049 million parameters and a computational cost of 0.060 GFLOPs, which is significantly lower than traditional convolutional neural networks and transformers used for image segmentation tasks. Despite its lightweight nature, the authors demonstrate that the UltraLight VM-UNet achieves competitive performance on three publicly available skin lesion segmentation datasets, outperforming several state-of-the-art lightweight models.

The success of the UltraLight VM-UNet can be attributed to the authors' in-depth analysis of the key factors influencing the parameters of the Vision Mamba module. By identifying the number of input channels as a critical factor affecting the parameter count, they were able to design the PVM Layer to process features in parallel while keeping the overall channel count constant, leading to a significant reduction in parameters without compromising performance.

\section{Our Models}

Inspired by \cite{U-Mamba} and \cite{ultralightvmunet}, we sought to implement both models on a different problem: tumor segmentation.

Our goal was to 1) maintain dice scores comparable to state-of-the-art (SOTA) methods such as those listed above, while 2) being smaller than those above.

We now go into what we did for each of our models.

\begin{figure}[!t]
    \centering
    \includegraphics[width=\columnwidth]{imgs/bars.png}
    \caption{Each model tested as a bar chart. The upper half is the Dice Scores, and the bottom half is the parameter count, where M means "million".}
    \label{fig:bars}
\end{figure}
\begin{figure}[!t]
    \centering
    \includegraphics[width=\columnwidth]{imgs/balls.png}
    \caption{A scatter plot of Dice Score vs model size in MegaBytes. The points are scaled to help represent size.}
    \label{fig:balls}
\end{figure}

\subsection{PP-UMamba}
This model incorporates pyramidal pooling, which is a strategy used in convolutional neural networks (CNNs) to capture context and incorporate multi-scale information \cite{zhao2017pyramid}. Pyramidal pooling involves parallel pooling operations at different scales, followed by concatenation of the resulting feature maps. This approach has been shown to improve the performance of CNNs in various computer vision tasks, including segmentation, by enabling the model to capture both local and global context.

In our PP-UMamba model, we integrated pyramidal pooling into the U-Mamba architecture, aiming to leverage the strengths of both the Mamba module and multi-scale feature extraction.

\subsection{ULMUNet\_v1}
For ULMUNet\_v1, the initial channel list was modified from [8, 16, 24, 32, 48, 64] to [16, 32, 64, 128, 256], increasing the number of channels at each step of the encoder and decoder. Additionally, the depth of the U-Net was reduced by one layer to keep the parameter count low while increasing the number of parameters in each step. The intention was to improve performance by increasing the capacity of the model while maintaining a reasonable parameter count.

\subsection{ULMUNet\_v2}
In ULMUNet\_v2, the number of parallel branches in the Parallel Vision Mamba (PVM) block was increased from 4 to 8. This modification directly decreases the parameter count, further proving the idea proposed in the UltraLight VM-UNet paper \cite{ultralightvmunet} that processing features in parallel with reduced channel counts can significantly reduce parameters while maintaining performance.

\subsection{ULMUNet\_v3}
ULMUNet\_v3 combines the modifications from ULMUNet\_v1 and ULMUNet\_v2. It increases the number of parallel branches in the PVM block to 8 and modifies the channel list to [16, 32, 64, 128, 256]. Additionally, the depth of the U-Net was reduced to 5 layers. This approach aims to balance the trade-off between model capacity and parameter efficiency.

\subsection{ULMUNet\_v4}
This model incorporates pyramidal pooling, similar to PP-UMamba, but within the UltraLight VM-UNet architecture. The goal was to leverage the benefits of multi-scale feature extraction while maintaining the parameter efficiency of the UltraLight VM-UNet.

\subsection{UL-PP}
The UL-PP (Ultra Light Pyramidal Pooling) model trained and performed well with 1.1M parameters. According to the provided information, incorporating pyramidal pooling on the UL network scored slightly worse than without pyramidal pooling but required approximately 400,000 more parameters.

\section{Results}
We now compare our models against ResNet18, ResNet50, standard U-Mamba with bottleneck, UNet, and UltraLight VM-UNet. The results are shown in Figures 1--2.

\section{Conclusion}

In this work, we explored the application of Structured State Space Sequence Models (SSMs), particularly the Mamba architecture, for tumor segmentation from brain MRI scans. The Mamba architecture leverages selective state spaces, enabling efficient and effective segmentation while maintaining linear scalability.

Our experiments and comparisons with state-of-the-art (SOTA) models demonstrate the potential of the Mamba-based approach in both model size and segmentation accuracy. The smallest model, ULVM\_Net\_v2, achieved a compelling dice score of 0.8243 with only 0.042M parameters, outperforming the larger ResNet50 model (42.9M params) with a dice score of 0.8672.

While the traditional U-Net model achieved the highest dice score of 0.8929, our PP-UMamba model closely followed with a dice score of 0.8867. Notably, our Mamba-based UMambaBot\_PP and UMambaBot models outperformed the Segformer transformer model, indicating the potential of Mamba-based approaches to surpass transformer-based models for this task.

We were able to make the ultra-light model smaller and more accurate on our task, further proving the potential of Mamba-based systems that prioritize compactness and efficiency. The selective state spaces and parallel processing strategies employed in our models enabled us to strike a balance between model size and segmentation accuracy.

Overall, this work contributes to the exploration of Structured State Space Sequence Models for medical image segmentation tasks. The Mamba architecture's ability to capture long-range dependencies while maintaining linear scalability positions it as a promising alternative to traditional convolutional and transformer-based approaches, especially in resource-constrained environments.

Future work could involve further refinements and optimizations of the Mamba-based models, as well as their application to other medical imaging modalities and segmentation tasks. Additionally, investigating the interpretability and robustness of these models could provide valuable insights for their deployment in real-world clinical settings.

@@

\bibitem{dosovitskiy2021image} A. Dosovitskiy et al., "An Image is Worth 16x16 Words: Transformers for Image Recognition at Scale," arXiv:2010.11929 [cs.CV], 2021.
\bibitem{ultralightvmunet} R. Wu, Y. Liu, P. Liang, and Q. Chang, “UltraLight VM-UNet: Parallel Vision Mamba Significantly Reduces Parameters for Skin Lesion Segmentation.” Accessed: Apr. 20, 2024. [Online]. Available: https://arxiv.org/pdf/2403.20035.pdf
\bibitem{zhao2017pyramid}
H. Zhao, J. Shi, X. Qi, X. Wang and J. Jia, "Pyramid Scene Parsing Network," 2017 IEEE Conference on Computer Vision and Pattern Recognition (CVPR), 2017, pp. 6230-6239, doi: 10.1109/CVPR.2017.660.
\end{thebibliography}
\end{document}
